Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features

### 2020-03-20 paper/paper.tex
Fixing typos

	The structure of this paper is as follows.
	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
	In section \ref{method} we consider non-parallel memory metrics collected by AIWC and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
	In section~\ref{discussion} we propose a new parallel spatial locality metric, implement it in the AIWC framework, and evaluate it over the matrix multiplication example from section~\ref{motivating example}.
	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.

@@

	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
	Global memory is typically located off-chip and accesses induce large delays.
	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory). 
	We see an improvement to 600 GFLOP/s is effective performance for order 2048 matrix multiplication.

@@

	The code is further optimized by improving the locality of reads from Matrix $B$.
	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
	The repeated reads of elements of matrix $B$ can be shared between threads in a work group by reading tiles from matrix $B$ into shared memory.
	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
	Performance is further improved to 614 GLOP/s after shared memory optimisation for Matrix B.

@@

	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
	Different ways of aggregating program measurements lead to different features of program execution being emphasised in the final metrics.
	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
	The order of sequential memory accesses performed by each thread, as well the relationship between work items in an OpenCL work group, are both vital in accurately characterizing parallel codes.

@@

	\textit{Summarize}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group. We then average the $n$-bits-dropped entropy summaries across thread groups to obtain the $n$-bits-dropped parallel spatial locality metrics for the kernel's execution. The theoretical maximum for an $n$-bits dropped parallel spatial locality value is $\log_2(group\_size)$. We divide the calculated average entropy values by $\log_2 (group\_size)$ to normalise the parallel spatial locality metric. This allows us to derive trends on the normalised metrics allowing comparison across separate OpenCL codes with different work distribution structures.

@@

	\section{Guidelines: Using the Parallel Spatial Locality Metric}

>>>>

	The structure of this paper is as follows.
	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
	In section \ref{method} we consider non-parallel memory metrics collected by AIWC and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
	In section~\ref{discussion} we propose a new parallel spatial locality metric, implement it in the AIWC framework, and evaluate it over the matrix multiplication example from section~\ref{motivating example}.
	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
	In section~\ref{guidelines} we summarize our findings and present guidelines for interpreting the new parallel spatial metric for OpenCL codes.
	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.

@@

	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
	Global memory is typically located off-chip and accesses induce large delays.
	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory). 
	We see an improvement to 600 GFLOP/s in effective performance for order 2048 matrix multiplication.

@@

	The code is further optimized by improving the locality of reads from Matrix $B$.
	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
	The repeated reads of elements of matrix $B$ can be shared between threads in a work group by reading tiles from matrix $B$ into shared memory.
	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
	Performance is further improved to 614 GLOP/s after this optimisation.

@@

	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
	Different ways of aggregating program measurements lead to different features of program execution being emphasized in the final metrics.
	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
	The order of sequential memory accesses performed by each thread, as well the relationship between work items in an OpenCL work group, are both vital in accurately characterizing parallel codes.

@@

	\textit{Summarize}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group. We then average the $n$-bits-dropped entropy summaries across thread groups to obtain the $n$-bits-dropped parallel spatial locality metrics for the kernel's execution. The theoretical maximum for an $n$-bits dropped parallel spatial locality value is $\log_2(group\_size)$. We divide the calculated average entropy values by $\log_2 (group\_size)$ to normalize the parallel spatial locality metric. This allows us to derive trends on the normalized metrics allowing comparison across separate OpenCL codes with different work distribution structures.

@@

	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}
